---
title: "R introduction: part II"
author: "Ming Chen"
date: "6/1/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```


# Functions

* Define a function:

```{r}
my_function = function(x, y, z, ...) {
  ## some code
  
  ## some other code
}
```


* Call the function

```{r}
my_function(x=VALUE, y=VALUE, z=VALUE)
```


* Example: define a function to extract all columns between two column names from a data frame. 

```{r}
## load data
pbsc = read.csv('PBSC-comma.txt', header = TRUE)

## we can get multiple columns by position
pbsc[2:4]

## get columns by names
get_columns_by_names = function(my_data_frame, first_name, last_name) {
  df_column_names = colnames(my_data_frame)
  first_position = which(df_column_names == first_name)
  last_position = which(df_column_names == last_name)

  my_data_frame[first_position:last_position]
}

## call the function
get_columns_by_names(pbsc, "A", "C")
get_columns_by_names(pbsc, "C", "A") # reversed
```


* Anonymous functions
    + Functions that don't have a name.
    + Usually used to create a closed environment for running some code.
    
```{r}
function(x, y, z, ...) {
  ## some code
  
  ## some other code
}

# will show you a use case in next section.
```


# Pipes and pipe operator

A pipe is a sequence of operations. It reflects the data and analysis flow.

The pipe operator `%>%` comes from the package **magrittr** by Stefan Milton Bache.

```{r}
# install.packages('magrittr')
# install.packages('dplyr')
library(dplyr)
pbsc %>% 
  head() %>% # the output from the previous operator will be the first argument in the next operator/function
  `[`(2:4) # yes! '[' is a function.

# the normal way to do this.
`[`(pbsc, 2:4) 

# use anonymous function in pipes
workshop_comment = 'I really do not like the RNASeq workshop!'
gsub('do not', 'do', workshop_comment)

workshop_comment %>%
  gsub('do not', 'do')  # this won't work because it takes workshop_comment as its first argument.


workshop_comment %>% 
  (function(comment) {
    gsub('do not', 'do', comment)
  })

```


# Data manipulation

We are going to learn five important functions from the **dplyr** package. These five functions allows you
to do a vast majority of data manipulation.

* `filter()`: filter rows
* `select()`: select variables by names
* `arrange()`: reorder the rows
* `mutate()`: create new variables by modifying existing variables
* `summarise()`: collapse many values down to a single summary

Here variables refer to data frame columns.

## Filter rows with `filter()`

`filter()` filters data frame by the logical values of expressions on columns. The first argument
is a data frame. The second and subsequent arguments are expressions.

```{r}
# load the library first
library(dplyr) 

pbsc
filter(pbsc, G > 25)
filter(pbsc, G > 25, A > 25) # it can takes multiple expressions
filter(pbsc, X.Base %in% 2:5 ) # get values from base 2 to 5

# what does the logical values of expressions mean?
pbsc$G > 25
pbsc$G >25 & pbsc$A > 25
pbsc$X.Base %in% 2:5

# filter NA rows
pbsc_na = pbsc
pbsc_na$phred = sample(c(20:60, rep(NA, 10)), nrow(pbsc_na), replace = TRUE)
pbsc_na

pbsc_na %>%
  filter(is.na(phred)) # select all rows that phred is NA

pbsc_na %>%
  filter(!is.na(phred)) # select all rows that phred is not NA
```


## Select columns by names with `select()`

```{r}
# select columns by names
select(pbsc, G, A)

# select all columns between G and C
select(pbsc, G:C)

# select all columns except those from G to C
select(pbsc, -(G:C))

# more cool stuff
iris

select(iris, starts_with('Sepal')) # names begin with 'Sepal'
select(iris, ends_with('Length')) # names end with 'Length'
select(iris, contains('al')) # names that contain 'al'
select(iris, matches("\\.")) # select by a regular expression
```


## Reorder rows with `arrange()`

```{r}
iris
arrange(iris, Petal.Length)
arrange(iris, Petal.Length, Petal.Width) # sort by multiple columns

arrange(iris, desc(Petal.Length)) # descending order
```


## Create new variables with `mutate()`

```{r}
mutate(pbsc, 
       GC = G + C,
       AT = A + G)


# refer to columns that you've just created
mutate(pbsc,
       GC = G + C,
       AT = A + G,
       GA_to_AT = GC/AT)

# if you only want to keep the new variables, use transmutate().
transmute(pbsc,
          GC = G + C,
          AT = A + G,
          GA_to_AT = GC/AT)

# a use case of combining multiple manipulating functions
#  goal: select rows that are GC biased
mutate(pbsc,
       GC = G + C,
       AT = A + G,
       GC_bias = GC > AT) %>%
  filter(GC_bias)
```


## Operating on subsets of a data frame with `summarise()`

`summarise()` is similar to `mutate()`. You operate on existing variables to new variables. But the new variables
have to be one single value.

```{r}
summarise(pbsc, GC = G + C) # this won't work because the new variable has length > 1

summarise(pbsc, average_G = mean(G)) # calculate the average of G content.

```

`summarise()` is usually used with `group_by()`. The `group_by()` converts a data frame to
a grouped table. So operations can be performed on by group.

```{r}
mtcars

grouped_mtcars = group_by(mtcars, cyl)
grouped_mtcars # what does the data look like now?

class(grouped_mtcars) # now it is a 'grouped_df' object.

summarise(grouped_mtcars, mean(mpg)) # calculate average mpg for each number of cyl.

# we can group the data by multiple variables.
group_by(mtcars, cyl, gear, carb) %>%
  summarise(mean(mpg), max(disp), min(wt))
```


