---
title: "R Intro I"
author: "Ming Chen"
date: "5/29/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

----------
# Get/Set working directory

```{r eval=FALSE}
getwd()
setwd('~/Desktop/rnaseq-workshop-2017')
```

----------
# Install/Uninstall and load packages

* Install package from CRAN

```{r eval=FALSE}
install.packages('ggplot2')
library('ggplot2')
require('ggplot2')
```

* Install package from local source file
    + google 'ggplot2 tar.gz'
    + https://cran.r-project.org/web/packages/ggplot2/index.html
```{r eval=FALSE}
remove.packages('ggplot2')
## restart R
library(ggplot2)
require('ggplot2')
install.packages('ggplot2_2.2.1.tar.gz', repos = NULL)
require('ggplot2')
```

* Install bioconductor packages
    + google 'deseq2 bioconductor'
    + https://www.bioconductor.org/packages/release/bioc/html/DESeq2.html

```{r eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("DESeq2")
require('DESeq2')
packageVersion('DESeq2')
```


* Install package from github
    + go to https://www.bioconductor.org/packages/release/bioc/html/DESeq2.html

```{r eval=FALSE}
remove.packages('DESeq2')
# restart R
require('DESeq2')

install.packages('devtools')
require('devtools')
install_github('Bioconductor-mirror/DESeq2')
install_github('Bioconductor-mirror/DESeq2', ref = 'release-3.3')
packageVersion('DESeq2')
## read this page if this doesn't work for you:
## https://stackoverflow.com/questions/29992066/rcpp-warning-directory-not-found-for-option-l-usr-local-cellar-gfortran-4-8/29993906#29993906
```

* Install package from command line
    + create an R script named 'install-r-packages.R' and put the following content into this file.
    + Open your terminal and run command: `RScript install-r-packages.R`

```{r eval=FALSE}
pkgs = c('plyr', 'dplyr')
install.packages(pkgs, repos = "http://cran.us.r-project.org")
```


# Get some help

* The `?` function.
* The `help()` function: `help(package='DESeq2')`
* **google**


# Data and Data Structures
## R's built-in datasets

```{r}
data(package='base')
data(package='ggplot2')
data('mpg', package='ggplot2')
mpg
```

## loading data

Two most common data loading functions: ***`read.csv()`*** and ***`read.table()`***.
    
```{r}
?read.csv
read.csv('PBSC-comma.txt')
read.csv('PBSC-tab.txt', sep = '\t')
read.csv('PBSC-comma.txt', header=FALSE)
```

```{r}
read.table('PBSC-tab.txt', header=TRUE) # does the header argument work?
read.table('PBSC-tab.txt', header=TRUE, comment.char = '')

# an example use case: load gtf file
read.table('Arabidopsis_thaliana.h20.gtf', header=FALSE, sep='\t', 
         comment.char = '#')
```


## variables

* Restrictions on variable names.
    + Technically amostly no restrictions on variable names
    + But **Most R programmers**
        * starts with lower case letters
        * separate words in names with underscores
        * only use lower case letters, numbers and underscores
```{r}
x = 1:10
x
my_data = 1:10
my_data
my_data_2 = 1:5
my_data_2

assign(x = 'my_data', value = 3:8)
my_data

# some variable names that you should probably never use.
assign(x = '2017', value = 1:10)
'2017'
get('2017')

assign(x = '#', value = 'this is my value')
get('#')
```
    
* Assign the imported data to a variable.
```{r}
pbsc = read.csv('PBSC-comma.txt', header = TRUE)
pbsc
```


## data structures

* Data frame
    + the most common data structure in R
    + has two dimensions
    
```{r}
is.data.frame(pbsc)

str(pbsc)
```


* Vectors
    + very basic data structure in R
    + two flavors: atomic vectors and lists.
```{r}
x = 1:10
x
is.vector(x)

# get a column from a data frame
pbsc$G
pbsc[, 1]
# a column from a data frame is an atomic vectors
is.vector(pbsc$G)
is.data.frame(pbsc$G)

# get a row from a data frame
pbsc[1, ]
# a row from a data frame is still a data frame.
is.vector(pbsc[1, ])
is.data.frame(pbsc[1, ])

pbsc[1,]$G
```


* atomic vectors vs. lists
    + atomic vectors are homogeneous. They can only have one type of values.
    + lists are homogeneous or heterogenous. 
    
```{r}
my_vector = c(1, 3, 5) # c() function combine values into a vector.
my_vector

my_v2 = c('a', 'b', 'c')
my_v2

my_v3 = c(1, 2, 'b') # automatically convert numbers to strings.
my_v3

my_list = list(a=1:3, b=c('a', 'b', 'd'))
my_list
my_list$a

# any types of data can be an element of a list.
my_l2 = list(a=1:3, b=c('a', 'b', 'd'), df=pbsc)
my_l2
my_l2$df
my_l2[1]
```


* matrix
    + two dimentional
    + **homogeneous**

```{r}
my_matrix = matrix(1:12, nrow=3, ncol=4)
my_matrix

my_matrix_2 = matrix(letters[1:12], nrow=3, ncol=4)
my_matrix_2

# convert a heterogeneous data frame to a homogeneous matrix
my_m3 = as.matrix(pbsc)
my_m3

# a column from a matrix is an vector
my_m3[, 3]
is.vector(my_m3[, 3])

# a row from a matrix is still an vector
my_m3[1, ]
is.vector(my_m3[1, ])
```

## Subsetting operators: `[[`, `$`, and `[`

* `$` is a shorthand for `[[`. It is useful when values are bound to character names. 
* `$` is invalid for atomic vectors.
* `[` is similar to `[[`. The difference is that `[[` can only return a single value, while `[[` can return multiple values.

```{r}
pbsc
```

* ***A data frame is also a list.***
    + *the columns in the data frame are elements in the list.*


```{r}
is.data.frame(pbsc)
is.list(pbsc)
```


### subsetting atomic vectors

```{r}
my_v1 = letters[1:10]
my_v1

# access single value
my_v1[1]
my_v1[[2]]

# access multiple values
my_v1[1:5]
my_v1[c(1,5,9)]
my_v1[c(1,3,3,3,1:10)] # get some elements repeatedly

my_v1[[1:5]] # this won't work

# access elements by names
names(my_v1) = LETTERS[1:10]
my_v1
my_v1['A']
my_v1[c('A','C','B')]

my_v1$A # this won't work!
```

```{r}
# A use case for accessing elements by names.

my_sequence = sample(c('g', 'c', 't', 'a'), 20, replace = TRUE)
my_sequence

# how to convert lower case bases to upper case bases?
my_template = c('G','C','T','A')
names(my_template) = c('g', 'c', 't', 'a')
my_template

my_template[my_sequence] # yeah!!
# you will see how useful this is when you use the ggplot2 package for data visualization.
```


### subsetting data frame

```{r}
# acess a single cell from data frame.
pbsc[1,2]
pbsc[[1,2]] # this will work since we are accessing single value.

# access multiple values
pbsc[1:5, ] # rows from 1 to 5
pbsc[, 1:2] # columns from 1 to 2
pbsc[1:5, 1:2] # subsetting by rows and columns

pbsc[[1:5, ]] # this won't work

# get one column from the data frame and remain the data structure.
pbsc[, 3]
is.vector(pbsc[, 3])

pbsc[, 3, drop=FALSE]
is.vector(pbsc[, 3, drop=FALSE])
is.data.frame(pbsc[, 3, drop=FALSE])

# elements in the index don't have to be unique.
pbsc[, c(1:4, 4, 4)]
nrow(pbsc)
pbsc[c(1:2, 2, 2, 3:55), ]

# use $
pbsc$A
pbsc$X.Base
```


### subsetting list

```{r}
my_l2 = list(a=1:3, b=c('a', 'b', 'd'), df=pbsc)
my_l2

# use $
my_l2$a
my_l2$df
```


```{r}
# use index to access multiple elements.
my_l2[1:3]
my_l2[c(1,1,1,1,2,2,3)]

my_l2[[1:3]] # this won't work!
```

**When `[` is applied to a list, it always returns a list. To access elements from an element within a list, we need the `[[` operator.**

```{r}
my_l2[1]
is.list(my_l2[1])

# how to get the second value from element 'a' in list 'my_l2'
my_l2[1][2] # this won't work!!
my_l2[1][1][1][1][1][1][1][1][1][1][1][1][1][1][1][1]

my_l2[[1]]
is.list(my_l2[[1]])
is.vector(my_l2[[1]])

my_l2[[1]][2] # this will return the second value from element 'a' in list 'my_l2'
my_l2[[3]]
my_l2[[3]]$G
```

### subsetting matrix

```{r}
pbsc_matrix = as.matrix(pbsc)
pbsc_matrix

# access columns
pbsc_matrix[, 2]
pbsc_matrix[, 2:3]

# access rows
pbsc_matrix[1:2, ]

# '$' operator does not apply to matrix.
pbsc_matrix$G # this won't work!!

# get single column or rows and remain the data structure.
pbsc_matrix[1, ]
pbsc_matrix[1, ,drop=FALSE]

pbsc_matrix[ , 2]
pbsc_matrix[ , 2,drop=FALSE]
```


## Factors